A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-08-28 doc/jcomp2012/jcomp2012.tex
Added reviewer comments to paper: Stuart, the Adjoint_Direct_Convergence.pdf figure needs axis labels.

    We present a novel synthetic-acceleration based Monte Carlo method
    for solving the equilibrium thermal radiation diffusion equation
    in three dimensions. The algorithm performance is compared against
    traditional solution techniques using a Marshak benchmark problem
    and a more complex multiple material problem. Our results show
    that not only can our Monte Carlo method be an effective solver
    for sparse matrix systems, but also that it performs competitively
    with deterministic methods including preconditioned Conjugate
    Gradient while producing numerically identical results. We also
    discuss various aspects of preconditioning the method and its
    general applicability to broader classes of problems.

@@

The purpose of this study is to demonstrate an efficient Monte Carlo
solution method for three dimensional (3D), time-dependent, discrete
systems. In this work, we extend our previous work in the following
manner:
\begin{itemize}
\item We develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems.
\item We apply this method to the 1T thermal radiation diffusion
  equation in three dimensions.
\item We compare numerical and performance results of the Monte Carlo
  method against traditional deterministic solution methods.
\end{itemize}

In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for
solving matrix systems. In \S~\ref{sec:iter-refin-monte} we discuss
our new synthetic-acceleration scheme and connect it with previous
work by Halton~\cite{halton_1994} and others~\cite{evans_2003}.  The
1T radiation diffusion equation and its discrete form will be derived
in \S~\ref{sec:discr-form-radi}.  We discuss using Monte Carlo
Synthetic-Accleration to solve the discrete radiation diffusion
equation in \S~\ref{sec:solut-radi-diff}.  Results from two 3D test
problems are given in \S~\ref{sec:results}.  We finish with
conclusions and ideas for future work in \S~\ref{sec:conclusions}.

@@

%%---------------------------------------------------------------------------%%
\subsection{Monte Carlo Matrix Solution Methods}
\label{sec:background}
To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius ($\rho$) of the iteration matrix ($\vH$) is
less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
  Series},
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation} 
Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
a series,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
With this knowledge in hand, we can write an iterative method that
solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the
Monte Carlo methods that are described in \S\S~\ref{sec:direct-method}
through \ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

@@

Now we consider a Monte Carlo method that can be used to estimate the
solution to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First,
rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
component of $\vx$,
\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\ &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}
The terms in the Neumann series can be interpreted as a series of
transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk
with $k$ events that initiates in state $i$,
\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\ &=
    \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}
Here, the weight on the $m^\text{th}$ step is denoted $W_m$, and each
random walk starts with unit weight. At every step, contributions to
each component of $X$ are generated only in the starting state of the
history, $i_0$, by the source term in the current state,
$b_{i_m}$. Therefore, for a given random walk permutation, each state
must be used as a starting value in order to calculate its component
of the expectation value (i.e. a state of size $N$ will have $N$
histories per random walk permutation). When there is a transition
between states, for example $i\rightarrow j$, the weight is multiplied
by the transition factor.
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
  \label{eq:weight}
\end{equation}
where $p_{ij}$ denotes the probability of transitioning from state $i$
to $j$. Then, the expected value of $X$ is
\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\ &=
    \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}
where $\nu$ denotes a particular random walk permutation.  Therefore,
the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased estimator
of the components of $\vx$ provided $\rho(\vH) < 1$.

@@

The following scheme will then converge in one iteration:
\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\ \vA \delta \vx^{l+1} =
    \vr^{l+1}\;,
    \label{eq:synthetic-residual-solve}\\ 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
To accelerate the fixed-point method and avoid inverting $\vA$ directly in
Eq.~(\ref{eq:synthetic-residual-solve}), we can instead use the above ideas to
create a new iterative scheme by using the Monte Carlo method to estimate the
solution error. With this in hand, the Fixed-Point Monte Carlo
Synthetic-Acceleration (MCSA) method can be defined as follows:
\begin{subequations}
  \label{eq:MCSA-iteration}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\  \vr^{l+1/2} = \vb -
    \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\     
     \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
     \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
\end{subequations}
The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that the
Monte Carlo solution only approximately inverts this operator.  Thus, we have
defined a scheme in which the initial estimate of $\vx$ in each iteration is
updated using a single fixed-point iteration\footnote{Although, a subspace
  method could be used instead of a stationary method here.}.  The residual is
calculated and used as a source to estimate the error, $\delta\vx^{l+1/2}$, by
solving Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the solution
vector, $\vx^{l+1}$.  The entire sequence is iterated to convergence based on
the following stopping criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

We choose to use the angle-energy integrated radiation intensity,
$\phi\equiv\phi(\ve{r},t)$
$[\text{GJ}\cdot\text{cm}^{-2}\cdot\text{ns}^{-1}]$, as the primary state
variable.  The intensity is defined as:
\begin{equation}
  \label{eq:lte-phi}
  \begin{split}
    \phi(\ve{r},t) &=
    \int_{4\pi}d\bOmega\int_{\nu}d\nu\:\psi(\ve{r},\nu,\hat{\bOmega},t)
    \\ &= \int_{\nu}d\nu\: 4\pi B(\nu,T) \\ &= acT^{4}\:,
  \end{split}
\end{equation}
where $\psi$ is the angular radiation intensity and $\nu$ is the radiation
frequency.  The Planck function (or Planckian) is defined as,
\begin{equation}
  B(\nu,T)=\frac{2h\nu^{3}}{c^{2}}\bigl(e^{\frac{h\nu}{kT}} -
  1\bigr)^{-1}\:,
\end{equation}
where $h$ is Planck's constant and $k$ is the Boltzmann constant.  Utilizing
this definition of $\phi$ we can write
\begin{equation}
  \label{eq:partial-T}
  \partial T = \frac{1}{4acT^{3}}\partial\phi\:.
\end{equation}
Inserting \eqref{eq:partial-T} into \eqref{eq:eq-diff-T} yields
\begin{equation}
  \label{eq:eq-diff}
  \Bigl(\frac{\rho\Cv}{4acT^{3}} + \frac{1}{c}\Bigr)\pder{\phi}{t} -
  \nabla\cdot D\grad{\phi} = Q\:,
\end{equation}
where $D$ is the diffusion coefficient, $D\equiv
D(\ve{r},T(\ve{r},t))\ [\text{cm}]$, and is defined as
\begin{equation}
  D = \frac{1}{3\ros}\:.
\end{equation}
Equation~\eqref{eq:eq-diff} is the model equation we will solve using the
Monte Carlo techniques presented in \S~\ref{sec:monte-carlo-matrix} and
\ref{sec:iter-refin-monte}.

@@

Applying backward-Euler time differencing to Eq.~(\ref{eq:eq-diff}) and
lagging the temperature-dependent coefficients at $t^n$ gives
\begin{equation}
  -\nabla\cdot \Dn\grad{\phi} + \sign\phi = \qn\:,
\end{equation}
where we have suppressed the $n+1$ indices on $\phi$ and
\begin{align}
  \sign &= \frac{\rho \Cv^n}{4ac(\Tn)^3\dt} + \frac{1}{c\dt}\:,\\ \qn
  &= \sign\phin + Q^n\:,\\ \phin &= ac(\Tn)^4\:.
\end{align}
Applying Fick's Law,
\begin{equation}
  \ve{F} = -D\grad{\phi}\:,
  \label{eq:fick}
\end{equation}
where $\ve{F}$ is the radiation flux, we define a flux-balance equation,
\begin{equation}
  \dive{F} + \sign\phi = \qn\:.
  \label{eq:flux-balance}
\end{equation}
Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
Fig.~\ref{fig:cell} gives
\begin{equation}
  \iiint\dive{F}\,d\ve{r} +
  \sign_{ijk}\phi_{ijk}V_{ijk}=\qn_{ijk}V_{ijk}\:,
  \label{eq:flux-balance-vol-integral}
\end{equation}
and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.
\begin{figure}[h!]
  \centerline{ \input{mesh_cell.pdftex_t}}
  \caption{Orthogonal 3D mesh cell.}
  \label{fig:cell}
\end{figure}
We now apply the divergence theorem,
\begin{equation}
  \begin{split}
    \iiint\dive{F}\,d\ve{r} &= \oint\hat{\ve{n}}\cdot\ve{F}\,dA\\ &=
    \sum_{f=1}^{6}\hat{\ve{n}}_f\cdot\ve{F}_f A_f\:,
  \end{split}
\end{equation}
to Eq.(\ref{eq:flux-balance-vol-integral}) yielding
\begin{equation}
  \begin{split}
    (F_{i+1/2} - F_{i-1/2})\Dj\Dk + (F_{j+1/2} &- F_{j-1/2})\Di\Dk +
    (F_{k+1/2}\\ &- F_{k-1/2})\Di\Dj + \sign_{ijk}\phi_{ijk}V_{ijk} =
    \qn_{ijk}V_{ijk}\:.
  \end{split}
  \label{eq:flux-balance-difference}
\end{equation}

@@

The discrete diffusion equation in (\ref{eq:discrete-diffusion}) can be
written in matrix-operator form as follows,
\begin{equation}
  \ve{D}\bphi = \ve{q}\:,
  \label{eq:diffusion-matrix}
\end{equation}
where $\ve{D}$ is a symmetric, positive-definite (SPD) operator.
Unfortunately, for many problems $\rho(\ve{D}) > 1$.  Therefore, we must
precondition Eq.~(\ref{eq:diffusion-matrix}) in order to use the MCSA
method. We apply left-preconditioning and write
\begin{equation}
  \ve{M}^{-1}\ve{D}\bphi = \ve{M}^{-1}\ve{q}\:.
  \label{eq:preconditioned-diffusion}
\end{equation}
A good choice for the preconditioner, $\ve{M}$, is to use Jacobi
preconditioning:
\begin{equation}
  \ve{M}=\diag(\ve{D})\:.
  \label{eq:preconditioner}
\end{equation}
Setting $\ve{M}$ equal to the diagonal elements of $\ve{D}$ accomplishes
several objectives:
\begin{itemize}
\item It reduces the spectral radius such that
  $\rho(\ve{M}^{-1}\ve{D}) < \rho(\ve{D})$ and
  $\rho(\ve{M}^{-1}\ve{D}) < 1$.  Thus, convergence is guaranteed and
  fewer iterations are required resulting in accelerated convergence.
\item The diagonal elements of the preconditioned iteration matrix are
  zero, $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no
  computations wasted for in-state transitions $i\rightarrow i$ during
  the random walk process.
\end{itemize}
The second item is quite important for Monte Carlo solutions of linear
systems, and this type of scaling should be performed regardless of any
additional preconditioner choices.

@@

\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
  \caption{ Two-dimensional plot of the temperature at 1000~ns on a
    cut-plane positioned at the mid-point of the $z$-axis.  The
    positions of the edit points where the time-evolution of the
    temperature is plotted in Fig.~\ref{fig:edit-points} are indicated
    by the black disks.}
  \label{fig:multi-mat_T}
\end{figure}

\begin{figure}[h!]
  \begin{center}
    \begin{tabular}{cc}
      \subfigure[Edit
        points]{\includegraphics[width=2in,clip]{points.pdf}} &
      \subfigure[Temperature]{\includegraphics[width=4in,clip]{multi_mat.pdf}}
    \end{tabular}
  \end{center}
  \caption{(a) Edit points and (b) time-evolution of the temperature
    at each point.  All points are centered in the $z$-axis.}
  \label{fig:edit-points}
\end{figure}

>>>>

    We present a novel synthetic-acceleration based Monte Carlo method for
    solving the equilibrium thermal radiation diffusion equation in three
    dimensions. The algorithm performance is compared against traditional
    solution techniques using a Marshak benchmark problem and a more complex
    multiple material problem. Our results show that not only can our Monte
    Carlo method be an effective solver for sparse matrix systems, but also
    that it performs competitively with deterministic methods including
    preconditioned conjugate gradient while producing numerically identical
    results. We also discuss various aspects of preconditioning the method and
    its general applicability to broader classes of problems.

@@

The purpose of this study is to demonstrate an efficient Monte Carlo solution
method for three dimensional (3D), time-dependent, discrete systems. In this
work, our previous investigations are extended in the following manner:
\begin{itemize}
\item we develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems;
\item we apply this method to the 1T thermal radiation diffusion
  equation in three dimensions;
\item we compare numerical and performance results of the Monte Carlo
  method against traditional deterministic solution methods.
\end{itemize}

In \S~\ref{sec:monte-carlo-matrix} we review Monte Carlo methods for solving
matrix systems, and we discuss our new synthetic-acceleration scheme and
connect it with previous work by Halton~\cite{halton_1994} and
others~\cite{evans_2003}.  The 1T radiation diffusion equation and its
discrete form will be derived in \S~\ref{sec:discr-form-radi}.  We discuss
using Monte Carlo Synthetic-Accleration to solve the discrete radiation
diffusion equation in \S~\ref{sec:solut-radi-diff}.  Results from two 3D test
problems are given in \S~\ref{sec:results}.  We finish with conclusions and
ideas for future work in \S~\ref{sec:conclusions}.

@@

%%---------------------------------------------------------------------------%%
\subsection{Monte Carlo Matrix Solution Methods}
\label{sec:background}

To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius ($\rho$) of the iteration matrix ($\vH$) is
less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
  Series},
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation} 
Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
a series,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
With this knowledge in hand, we can write an iterative method that
solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb,\vx^0 \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the Monte
Carlo methods that are described in \S\S~\ref{sec:direct-method} through
\ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

@@

Now we consider a Monte Carlo method that can be used to estimate the
solution to Eq.~(\ref{eq:Ax=b})~\cite{hammersley_1964}.  First,
rewrite Eq.~(\ref{eq:neumann_series}) in order to calculate a
component of $\vx$,
\begin{equation}
  \begin{split}
    x_i &= (\vb)_i + (\vH\vb)_i + (\vH^2\vb)_i + (\vH^3\vb)_i +
    \ldots\\ &= \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}h_{i,i_1}h_{i_1,i_2}\ldots h_{i_{k-1},i_k}b_{i_k}\:.
  \end{split}
  \label{eq:neumann_series_i}
\end{equation}
The terms in the Neumann series can be interpreted as a series of
transitions from $i_{m-1}\rightarrow i_m$ that can be simulated by a
random walk.  Let $X$ be a random variable sampled from a random walk
with $k$ events that initiates in state $i$,
\begin{equation}
  \begin{split}
    X(i_0 = i) &= \sum_{m=0}^{k}W_m b_{i_m}\\ &=
    \sum_{m=0}^{k}w_{i,i_1}w_{i_1,i_2}\ldots w_{i_{m-1},i_m}b_{i_m}\:.
  \end{split}
  \label{eq:estimator_xi}
\end{equation}
Here, the weight on the $m^\text{th}$ step is denoted $W_m$, and each
random walk starts with unit weight. At every step, contributions to
each component of $X$ are generated only in the starting state of the
history, $i_0$, by the source term in the current state,
$b_{i_m}$. Therefore, for a given random walk permutation, each state
must be used as a starting value in order to calculate its component
of the expectation value (i.e. a state of size $N$ will have $N$
histories per random walk permutation). When there is a transition
between states, for example $i\rightarrow j$, the weight is multiplied
by the transition factor
\begin{equation}
  w_{ij} = \frac{h_{ij}}{p_{ij}}\:,
  \label{eq:weight}
\end{equation}
where $p_{ij}$ denotes the probability of transitioning from state $i$
to $j$. Then, the expected value of $X$ is
\begin{equation}
  \begin{split}
    E[X(i_0=i)] &= \sum_{\nu}P_\nu X_\nu\\ &=
    \sum_{k=0}^{\infty}\sum_{i_1}^{N}\sum_{i_2}^{N}\ldots
    \sum_{i_k}^{N}p_{i,i_1}p_{i_1,i_2}\ldots p_{i_{k-1},i_k}
    w_{i,i_1}w_{i_1,i_2}\dots w_{i_{k-1},i_k}b_{i_k}\\ &= x_i\:,
  \end{split}
  \label{eq:expectation_xi}
\end{equation}
where $\nu$ denotes a particular random walk permutation.  Therefore,
the estimator in Eq.~(\ref{eq:estimator_xi}) is an unbiased estimator
of the components of $\vx$ provided $\rho(\vH) < 1$.

@@

The following scheme will then converge in one iteration:
\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\ \vA \delta \vx^{l+1} =
    \vr^{l+1}\;,
    \label{eq:synthetic-residual-solve}\\ 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
To accelerate the fixed-point method and avoid inverting $\vA$ directly in
Eq.~(\ref{eq:synthetic-residual-solve}), we can instead use the above ideas to
create a new iterative scheme that applies the Monte Carlo method to estimate
the solution error. With this in hand, the Fixed-Point Monte Carlo
Synthetic-Acceleration (MCSA) method can be defined as follows:
\begin{subequations}
  \label{eq:MCSA-iteration}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\  \vr^{l+1/2} = \vb -
    \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\     
     \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
     \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
\end{subequations}
The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that the
Monte Carlo solution only approximately inverts this operator.  Thus, we have
defined a scheme in which the initial estimate of $\vx$ in each iteration is
updated using a single fixed-point iteration\footnote{Although, a subspace
  method could be used instead of a stationary method here.}.  The residual is
calculated and used as a source to estimate the error, $\delta\vx^{l+1/2}$, by
solving Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the solution
vector, $\vx^{l+1}$.  The entire sequence is iterated to convergence based on
the following stopping criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

We choose to use the angle-energy integrated radiation intensity,
$\phi\equiv\phi(\ve{r},t)$
$[\text{GJ}\cdot\text{cm}^{-2}\cdot\text{ns}^{-1}]$, as the primary state
variable.  The intensity is defined as:
\begin{equation}
  \label{eq:lte-phi}
  \begin{split}
    \phi(\ve{r},t) &=
    \int_{4\pi}d\bOmega\int_{\nu}d\nu\:\psi(\ve{r},\nu,\hat{\bOmega},t)
    \\ &= \int_{\nu}d\nu\: 4\pi B(\nu,T) \\ &= acT^{4}\:,
  \end{split}
\end{equation}
where $\psi$ is the angular radiation intensity and $\nu$ is the radiation
frequency.  The Planck function (or Planckian) is defined as,
\begin{equation}
  B(\nu,T)=\frac{2h\nu^{3}}{c^{2}}\bigl(e^{\frac{h\nu}{kT}} -
  1\bigr)^{-1}\:,
\end{equation}
where $h$ is Planck's constant and $k$ is the Boltzmann constant.  Utilizing
this definition of $\phi$ we can write
\begin{equation}
  \label{eq:partial-T}
  \pder{T}{t} = \frac{1}{4acT^{3}}\pder{\phi}{t}\:.
\end{equation}
Inserting \eqref{eq:partial-T} into \eqref{eq:eq-diff-T} yields
\begin{equation}
  \label{eq:eq-diff}
  \Bigl(\frac{\rho\Cv}{4acT^{3}} + \frac{1}{c}\Bigr)\pder{\phi}{t} -
  \nabla\cdot D\grad{\phi} = Q\:,
\end{equation}
where $D$ is the diffusion coefficient, $D\equiv
D(\ve{r},T(\ve{r},t))\ [\text{cm}]$, and is defined as
\begin{equation}
  D = \frac{1}{3\ros}\:.
\end{equation}
Equation~\eqref{eq:eq-diff} is the model equation we will solve using the
Monte Carlo techniques presented in \S~\ref{sec:monte-carlo-matrix} and
\ref{sec:iter-refin-monte}.

@@

Applying backward-Euler time differencing to Eq.~(\ref{eq:eq-diff}) and
lagging the temperature-dependent coefficients at $t^n$ gives
\begin{equation}
  -\nabla\cdot \Dn\grad{\phi} + \sign\phi = \qn\:,
\end{equation}
where we have suppressed the $n+1$ indices on $\phi$ and
\begin{align}
  \sign &= \frac{\rho \Cv^n}{4ac(\Tn)^3\dt} + \frac{1}{c\dt}\:,\\ \qn
  &= \sign\phin + Q^n\:,\\ \phin &= ac(\Tn)^4\:.
\end{align}
Applying Fick's Law,
\begin{equation}
  \ve{F} = -D\grad{\phi}\:,
  \label{eq:fick}
\end{equation}
where $\ve{F}$ is the radiation flux, we define a flux-balance equation,
\begin{equation}
  \dive{F} + \sign\phi = \qn\:.
  \label{eq:flux-balance}
\end{equation}
Integrating Eq.~(\ref{eq:flux-balance}) over the cell illustrated in
Fig.~\ref{fig:cell} gives
\begin{equation}
  \iiint\dive{F}\,d\ve{r} +
  \sign_{ijk}\phi_{ijk}V_{ijk}=\qn_{ijk}V_{ijk}\:,
  \label{eq:flux-balance-vol-integral}
\end{equation}
and $V_{ijk} = \Delta_i\Delta_j\Delta_k$.
\begin{figure}[h!]
  \centerline{ \input{mesh_cell.pdftex_t}}
  \caption{Orthogonal 3D mesh cell.}
  \label{fig:cell}
\end{figure}
We now apply the divergence theorem,
\begin{equation}
  \begin{split}
    \iiint\dive{F}\,d\ve{r} &= \oint\hat{\ve{n}}\cdot\ve{F}\,dA\\ &=
    \sum_{f=1}^{6}\hat{\ve{n}}_f\cdot\ve{F}_f A_f\:,
  \end{split}
\end{equation}
to Eq.(\ref{eq:flux-balance-vol-integral}) yielding
\begin{equation}
  \begin{split}
    (F_{i+1/2} - F_{i-1/2})\Dj\Dk &+ (F_{j+1/2} - F_{j-1/2})\Di\Dk +
    (F_{k+1/2}\\ &- F_{k-1/2})\Di\Dj + \sign_{ijk}\phi_{ijk}V_{ijk} =
    \qn_{ijk}V_{ijk}\:.
  \end{split}
  \label{eq:flux-balance-difference}
\end{equation}

@@

The discrete diffusion equation in (\ref{eq:discrete-diffusion}) can be
written in matrix-operator form as follows,
\begin{equation}
  \ve{D}\bphi = \ve{q}\:,
  \label{eq:diffusion-matrix}
\end{equation}
where $\ve{D}$ is a symmetric, positive-definite (SPD) operator.
Unfortunately, for many problems $\rho(\ve{D}) > 1$.  Therefore, we must
precondition Eq.~(\ref{eq:diffusion-matrix}) in order to use the MCSA
method. We apply left-preconditioning and write
\begin{equation}
  \ve{M}^{-1}\ve{D}\bphi = \ve{M}^{-1}\ve{q}\:.
  \label{eq:preconditioned-diffusion}
\end{equation}
A good choice for the preconditioner, $\ve{M}$, is to use Jacobi
preconditioning:
\begin{equation}
  \ve{M}=\diag(\ve{D})\:.
  \label{eq:preconditioner}
\end{equation}
Setting $\ve{M}$ equal to the diagonal elements of $\ve{D}$ accomplishes
several objectives:
\begin{itemize}
\item For diagonally dominant systems like Eq.~(\ref{eq:diffusion-matrix}), it
  reduces the spectral radius such that $\rho(\ve{M}^{-1}\ve{D}) <
  \rho(\ve{D})$ and $\rho(\ve{M}^{-1}\ve{D}) < 1$ \cite{golub}.  Thus,
  convergence is guaranteed and fewer iterations are required resulting in
  accelerated convergence.
\item The diagonal elements of the preconditioned iteration matrix are
  zero, $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no
  computations wasted for in-state transitions $i\rightarrow i$ during
  the random walk process.
\end{itemize}
The second item is quite important for Monte Carlo solutions of linear
systems, and this type of scaling should be performed regardless of any
additional preconditioner choices.

@@

\begin{figure}[h!]
  \centerline{ \includegraphics[width=5in,clip]{T_multi_mat.pdf}}
  \caption{ Two-dimensional plot of the temperature in keV at 1000~ns on a
    cut-plane positioned at the mid-point of the $z$-axis.  The positions of
    the edit points where the time-evolution of the temperature is plotted in
    Fig.~\ref{fig:edit-points} are indicated by the black disks.}
  \label{fig:multi-mat_T}
\end{figure}

\begin{figure}[h!]
  \begin{center}
    \includegraphics[width=5in,clip]{edit_points}
  \end{center}
  \caption{(a) Edit points and (b) time-evolution of the temperature at each
    point.  All points are centered in the $z$-axis.}
  \label{fig:edit-points}
\end{figure}
